Appendix 1: Quantifying Drought

### 2020-11-19 analysis/supplementary-materials/appendix_SPEI.Rmd
work on plot to compare data sources. Added EMBRAPA station data

```{r read-data}
# Precip Sources
xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
trmm <-

# Already calculated SPEI  
gspei <-
  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
             delim = ";") %>% 
  rename(date = "days since 1900-1-1", g_spei = spei)

@@

@@

-   Daily precipitation grids for South America [@liebmannDailyPrecipitationGrids2005] <!--# Not sure this one even makes sense to use, as it doesn't include ETO or temperature data, so I'd have to get that from Xavier or another source. -->

    -   Resolution 1º x 1º

    -   Includes only precipitation

    -   PET calculated by \_\_\_ method using \_\_\_ data

    -   Source: <!--# number --> ground-based weather stations across South America

    -   Date range: 1987--2012

@@

```{r sa-spei}
# sa # I'm not sure this one is worth adding to the comparison since there is no temperature or ETO data associated with it.

@@

# For Cronians talk
Just gonna make a simple-ish plot (lol) with major droughts labeled

```{r}
xa_df <-
  xa_spei %>% 
  group_by(yearmonth) %>% 
  #Summarize across grid cells
  summarize(`Xavier et al.` = mean(xa_spei))

comb <- left_join(xa_df, gspei %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)) %>% 
  rename(`Global SPEI` = g_spei) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01")))
```

@@

Try labeling droughts

```{r}
library(gghighlight)
p + gghighlight(year(yearmonth) ==2010, use_direct_label = FALSE)
```

>>>>

```{r read-data}
# Precip Sources
xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
# trmm <-

# Already calculated SPEI  
gspei <-
  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
             delim = ";") %>% 
  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)

@@

### Meterological stations

-   EMBRAPA stations

    -   Source:

    -   Date Range:

    -   PET calcualted by ... method

@@

@@

```{r embrapa-spei}
rpde <-
  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv")) %>% 
  select(date = Data,
         precip = Precipitacao,
         eto = Evatotranspiracao,
  )
manaus <-
  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Manaus_1980-01-01_2016-12-31.csv"))%>% 
  select(date = Data,
         precip = Precipitacao,
         eto = Evatotranspiracao,
  )

embrapa_mon <-
  bind_rows("rpde" = rpde, "manaus" = manaus, .id = "station") %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  group_by(yearmonth, station) %>% 
  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
  as_tsibble(index = yearmonth, key = station)

embrapa_spei <-
  embrapa_mon %>% 
  mutate(cb = precip - eto) %>% 
  group_by(station) %>% 
  group_nest() %>% 
  mutate(st_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(st_spei, data)) %>% 
  select(-cb)

embrapa_wide <-
  embrapa_spei %>% 
  select(yearmonth, station, st_spei) %>% 
  pivot_wider(names_from= station, values_from = st_spei, names_prefix = "spei_") %>% 
  rowwise() %>% 
  mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
  ungroup()
```

@@

# Plot

Create a plot comparing data sources with known drought years labeled.

## Combine into single tidy data frame

```{r}
# take average across site
xa_df <-
  xa_spei %>% 
  group_by(yearmonth) %>% 
  #Summarize across grid cells
  summarize(xa_spei = mean(xa_spei))

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  rename(`Global SPEI` = g_spei,
         `Xavier et al.` = xa_spei,
         `EMBRAPA` = st_spei) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01")))
```

Annotate worst droughts.  
Expected:

- 2004 & 2006: weak ENSO
- 2002 & 2009: moderate ENSO
- 1997: severe ENSO
- 2005 & 2010: severe SST

```{r}
comb <- 
  comb %>% 
  mutate(year = year(yearmonth)) %>% 
  group_by(year) %>% 
  mutate(min_mon = spei == min(spei)) %>% 
  mutate(drought_label = case_when(
    year %in% c(2004, 2006) & min_mon ~ "weak ENSO",
    year %in% c(2002, 2009) & min_mon ~ "moderate ENSO",
    year == 1997 & min_mon ~ "severe ENSO",
    year %in% c(2005, 2010) & min_mon ~ "severe SST"
    ))
```

```{r}
labs <- comb %>% 
  filter(!is.na(drought_label)) %>% 
  group_by(drought_label) %>% 
  mutate(xend = mean(yearmonth))

p <-
  ggplot(comb, aes(x = yearmonth, y = spei)) +
  geom_line(aes(linetype = dataset), size = 0.5) +
  # geom_label_repel(data = comb %>% filter(!is.na(drought_label)),
  #                  aes(label = drought_label, y = -2.5))
  geom_segment(data = labs,
               aes(xend = yearmonth, yend = spei -1)) +
  geom_label(data = labs,
             aes(x = yearmonth, y = spei-1, label = year, fill = drought_label)) +
  scale_x_yearmonth(date_breaks = "2 years") +
  scale_color_brewer(type = "qual", palette = 2) +
  coord_cartesian(xlim = c(yearmonth(ymd("1997-01-01")), yearmonth(ymd("2010-11-01")))) +
  theme_bw()
annotate_spei(p)
```

For Cronian's talk:

@@
